Linking watershed nutrient loading to estuary water quality with semiparametric models

### 2023-06-23 Schramm-2023-06-PeerJ/Schramm-2023-06-PeerJ.tex
peerj submission files

\begin{abstract}
Evaluating estuary water quality responses to reductions (or increases)
in nutrient loading attributed to on the ground management actions can
be challenging due to the strong influence of environmental drivers on
nutrient loads and non-linear relationships. This study applied
semiparametric approaches to calculate watershed nutrient loads and
assess responses in estuary water quality to seasonally-adjusted
freshwater inflow and flow-adjusted nutrient loads in Lavaca Bay, Texas.
Lavaca Bay is a secondary embayment on the Texas coast displaying early
potential for eutrophication and water quality degredation. Use of
flow-adjusted nutrient loads allowed the study to evaluate the response
in water quality to changes in nutrient loads driven by non-natural
variation. Cross-validation indicated that, despite data constraints,
semiparametric models performed well at nutrient load prediction. Based
on these models, delivered annual nutrient loads varied substantially
from year to year. In contrast, minimal changes in flow-normalized loads
indicate that nutrient loadings were driven by natural variation in
precipitation and runoff as opposed to changes in management of nonpoint
sources. Models indicated no evidence of long-term changes in dissolved
oxygen or chlorophyll-\emph{a} within Lavaca Bay. However, site specific
long-term increases in both organic and inorganic nitrogen are causes of
concern as potential causes of future eutrophication. Further analysis
found freshwater inflow was a strong driver of nutrient and
chlorophyll-\emph{a} concentrations within Lavaca Bay but there was no
evidence that changes in watershed nutrient loading explained additional
variation in dissolved oxygen or chlorophyll-\emph{a} concentrations. In
addition to providing a baseline assessment of watershed nutrient
loading and water quality responses in the Lavaca Bay watershed, this
study provides methodological support for the use of semiparametric
methods in load regression models and estuary assessments.
% Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text.
\end{abstract}

@@

Similar to many coastal areas globally, the coastal watersheds along the
Texas Gulf coast are facing pressures from increasing population,
increases in point source and non-point source pollution and alterations
to freshwater flows that degrade water quality in downstream estuaries
\autocite{bricker_effects_2008,kennicuttWaterQualityGulf2017,bugica_water_2020}.
Despite these escalating pressures, national scale assessments have
classified coastal estuaries in Texas as moderate or low risk for
eutrophic conditions \autocite{bricker_effects_2008}. However, a suite
of recent studies indicates that estuary water quality dynamics in both
agriculturally dominated and urban watersheds within Texas are
expressing conditions that are increasingly conducive to algal blooms
and eutrophication
\autocite{wetzWaterQualityDynamics2016,wetz_exceptionally_2017,bugica_water_2020,chinPhytoplanktonBiomassCommunity2022}.
With identification of several localized areas of estuary water quality
concern along the Texas coast \autocite{bugica_water_2020}, localized
studies are being prioritized to better inform management actions.

This goal of this project is to assess watershed nutrient loading and
the resulting water quality responses in Lavaca Bay, Texas. Lavaca Bay
is a secondary bay in the larger Matagorda Bay system located roughly
halfway between Houston, Texas and Corpus Christi, Texas. Lavaca Bay
faces substantial challenges associated with legacy contamination but
general water quality parameters such as dissolved oxygen (DO),
nutrients, and biological parameters have been well within state water
quality standards. Despite largely meeting statewide water quality
standards, there have been concerning declines in abundance, biomass,
and diversity of benthic fauna in Lavaca Bay
\autocite{beserespollackLongtermTrendsResponse2011}. These declines are
partially attributed to reductions in freshwater inflow and changes in
estuary salinity and are indicative of an already stressed system
\autocite{beserespollackLongtermTrendsResponse2011,palmerImpactsDroughtsLow2015,montagnaAssessmentRelationshipFreshwater2020}.
More recently, significant linear increases in total phosphorus (TP),
orthophosphate, total Kjeldahl nitrogen (TKN), and chlorophyll-\emph{a}
concentrations were identified at monitoring sites within Lavaca Bay
\autocite{bugica_water_2020}. Although long-term changes in DO
concentrations have not been identified, the trends in nutrient
concentrations are concerning due to the role of nitrogen as a limiting
factor for primary production in many Texas estuaries
\autocite{gardnerNitrogenFixationDissimilatory2006,houTransformationFateNitrate2012,doradoUnderstandingInteractionsFreshwater2015,paudelRelationshipSuspendedSolids2019,wetz_exceptionally_2017}
and the ramifications that changes in nitrogen loadings could have for
productivity and eutrophication in Lavaca Bay.

There are ongoing efforts between local, state, and federal agencies to
address water quality impairments in the freshwater portions of the
Lavaca Bay watershed
\autocite{jainTechnicalSupportDocument2021,schrammLavacaRiverWatershed2018,bertholdDirectMailingEducation2021}.
However, at a statewide scale, these approaches have shown limited
success and emphasize a need for improved efforts at assessing and
linking management actions with downstream water quality to identify and
replicate effective management actions across the state
\autocite{schrammTotalMaximumDaily2022}. The identification and
communication of changes and trends in water quality is complicated by
the fact that trends are often non-linear and confounded by
precipitation and runoff that hinder traditional analysis
\autocite{wazniakLinkingWaterQuality2007,lloydMethodsDetectingChange2014}.
The development and application of flexible statistical methods such as
Weighted Regressions on Time, Discharge and Season
\autocite[WRTDS,][]{hirschWeightedRegressionsTime2010} and Generalized
Additive Models \autocite[GAMs,][]{woodFastStableRestricted2011} have
provided effective tools for researchers to quantify and communicate
non-linear changes in river and estuary pollutant loadings.

WRTDS calculates a time series of in-stream concentrations or loads
(daily, monthly, or annually) and flow-normalized estimates of
concentrations and loads using locally weighted regresssion for unique
combinations of time, discharge, and season. WRTDS has been widely used
to assess and identify trends in riverine nutrients
\autocite{oelsner_recent_2019,stackpooleLongTermMississippi2021},
chlorides \autocite{stetsIncreasingChlorideRivers2018}, and other
pollutants of concern \autocite{shodaWaterqualityTrendsRivers2019}.
WRTDS has also been succesfully adapted to assess trends in estuarine
water quality concentrations \autocite{beckFourDecadesWater2018}.

While WRTDS is a statistical approach developed specifically for water
quality applications, GAMs are a broadly applicable statistical method.
GAMs are a semiparametric extension of generalized linear models where
the linear predictor is represented as the sum of multiple unknown
smooth functions and parametric linear predictors
\autocite{woodFastStableRestricted2011}. Although the underlying
parameter estimation procedure of GAMs are substantially different than
WRTDS, both the functional form and results have been demonstrated to be
similar when assessing nutrient concentration trends
\autocite{beckNumericalQualitativeContrasts2017}. Water quality
applications of GAMs have included river and catchment nutrient
concentration and load estimation
\autocite{wangLoadEstimationUncertainties2011,kroonRiverLoadsSuspended2012,kuhnertQuantifyingTotalSuspended2012,robson_prediction_2015-1,hagemannEstimatingNutrientOrganic2016,mcdowell_implications_2021,biagi_novel_2022},
and assessment of temporal trends of nutrients
\autocite{beckNumericalQualitativeContrasts2017,murphyGeneralizedAdditiveModel2019},
phytoplankton \autocite{bergbuschUnexpectedShiftPhytoplankton2021}, and
cyanobacteria \autocite{hayesEffectsLakeWarming2020}. Recently GAMs have
also been used to link water quality responses in receiving water bodies
to changes in nonpoint source nutrient inputs
\autocite{murphyNutrientImprovementsChesapeake2022}. For a substantial
discussion on the differences (and similarities) between GAMs and WRTDS
for water quality applications reders are referred to
\textcite{beckNumericalQualitativeContrasts2017}.

@@

{\centering \includegraphics[width=5.2in,]{../fig1}

@@

Water quality data for the two freshwater and three estuary locations
were obtained from the Texas Commission on Environmental Quality (TCEQ)
Surface Water Quality Monitoring Information System. Data submitted
through the system are required to be collected under Quality Assurance
Project Plans and lab method procedures outlined by the TCEQ's
procedures manual to ensure consistent collection and laboratory methods
are applied between samples collected by different entities and under
different projects. All sites had varying lengths of and availability of
data. For freshwater locations, TP from January 2000 through December
2020 and nitrate-nitrogen (NO\textsubscript{3}) data from January 2005
through December 2020 were downloaded (Table~\ref{tab:table1}). Less
than 5-years of total nitrogen and TKN concentration data were available
at the freshwater sites and deemed insufficient to develop load
estimation models
\autocite{horowitzEvaluationSedimentRating2003,snelderEstimationCatchmentNutrient2017}.
The three estuary sites included an upper Lavaca Bay site near the
outlet of the Lavaca River system (TCEQ-13563), a mid-Lavaca Bay site
(TCEQ-13383), and the lower Lavaca Bay site near the mouth of the Bay
(TCEQ-13384). For estuary locations, we obtained data for TP,
Nitrite+Nitrate (NO\textsubscript{\emph{X}}), TKN, chlorophyll-\emph{a},
and DO concentrations from January 2005 through December 2020
(Table~\ref{tab:table2}).

@@

\begin{align}
g(\mu) &= \alpha + f_1(ddate) + f_2(yday) + f_3(log1p(Q)) + f_4(ma) + f_5(fa), \nonumber \\
y &\sim \mathcal{N}(\mu,\,\sigma^{2})
\label{eq1}
\end{align}

where \(\mu\) is the conditional expected NO\textsubscript{3} or TP
concentration, \(g()\) is the log-link, \(\alpha\) is the intercept,
\(f_n()\) are smoothing functions. \(y\) is the response variable
(NO\textsubscript{3} or TP concentration) modeled as normally
distributed with mean \(\mu\) and standard deviation \(\sigma\).
\emph{ddate} is the date converted to decimal notation, \emph{yday} is
numeric day of year (1-366), and \emph{log1p(Q)} is the natural log of
mean daily streamflow plus 1.

@@

Daily loads were estimated as the predicted concentration multiplied by
the daily streamflow. For the Navidad River (USGS-08164525) site, daily
loads at the dam were calculated from the discrete daily concentration
at the discharge point of the lake and corresponding reported daily
discharge from the dam. Flow-normalized loads were estimated similar to
WRTDS by setting flow-based covariates on each day of the year equal to
each of the historical values for that day of the year over the study
period \autocite{hirschWeightedRegressionsTime2010}. The flow-normalized
estimate was calculated as the mean of all the predictions for each day
considering all possible flow values. Standard deviations and 90\%
credible intervals were obtained by drawing samples from the
multivariate normal posterior distribution of the fitted GAM
\autocite{woodConfidenceIntervalsGeneralized2006,marraCoveragePropertiesConfidence2012,mcdowell_implications_2021}.
GAM performance was evaluated using repeated 5-fold cross validation
\autocite{burmanComparativeStudyOrdinary1989} and average Nash-Sutcliffe
Efficiency (NSE), Pearson sample correlation (\emph{r}) and percent bias
(PBIAS) metrics across folds were calculated for each model.

@@

Because streamflow and nutrient loads are tightly correlated, freshwater
inflow can mask signals due to chnages in nutrient loads alone.
Following the methodology implemented by
\textcite{murphyNutrientImprovementsChesapeake2022}, both freshwater
inflow and nutrient loads were prepossessed to account for season and
streamflow respectively. Raw freshwater inflow values were replaced by
seasonally adjusted log transformed inflow obtained from the residuals
of a GAM model fit between season(day of year) and log transformed daily
freshwater inflow. Raw nutrient loads were replaced with flow-adjusted
values obtained from the residuals of a GAM model relating log
transformed NO\textsubscript{3} or TP loads to log transformed daily
inflow. Response residuals from the respective GAM models were used as
\emph{Q} and \emph{Load} in Eq.~\ref{eq6} and Eq.~\ref{eq7}.

@@

Predictive performance of nutrient loads ranged from ``satisfactory'' to
``very good'' based on standardized evaluation metrics of NSE, \emph{r},
and PBIAS \autocite{moriasiHydrologicWaterQuality2015} calculated using
5-fold cross validation. Median goodness-of-fit metrics for
NO\textsubscript{3} models in the Lavaca River were 0.34 NSE, 0.70
\emph{r}, and 2.00 PBIAS. Navidad River NO\textsubscript{3} models
appeared to perform slightly better with 0.48 NSE and 0.87 \emph{r} but
with higher bias at 10.90 PBIAS. Generally, TP models performed better
than NO\textsubscript{3} models. Median goodness-of-fit metrics for TP
in the Lavaca River were 0.81 NSE, 0.93 \emph{r}, and -7.20 PBIAS.
Navidad River TP models has similar performance with 0.91 NSE, 0.99
\emph{r}, and -3.30 PBIAS. Density plots of metrics show similar
distribution of values between sites for the same parameter, with the
exception \emph{r} values for NO\textsubscript{3} loads where Lavaca
River had a much larger variance in values compared to the Navidad River
(Fig.~\ref{fig:fig2}). TP GAMS had higher average NSE and \emph{r}
values and lower variance in metric values compared to
NO\textsubscript{3}.

@@

{\centering \includegraphics[width=1\linewidth,]{Schramm-2023-06-PeerJ_files/figure-latex/fig2-1}

@@

{\centering \includegraphics[width=1\linewidth,]{Schramm-2023-06-PeerJ_files/figure-latex/fig3-1}

@@

Aggregated across both sites, the mean annual NO\textsubscript{3} load
2005 through 2020 was 205,405 kg (126,867 kg - 341,569 kg, 90\% CI).
Annual NO\textsubscript{3} loads ranged from 12,574 kg in 2011 to
794,510 kg in 2007. Total annual TP loads ranged from 7,839 kg in 2011
to 595,075 kg in 2007. Mean annual TP loading from 2005 through 2020 was
182,673 kg (152,227 kg - 219,310 kg, 90\% CI). On average, the Navidad
River accounted for 68\% of NO\textsubscript{3} loads and 59\% of TP
loads from 2005 through 2020. However, during periods of extreme drought
the Lavaca River became the primary source of nutrient loading in the
watershed with the Navidad River only accounting for 15\% and 25\% of
NO\textsubscript{3} and TP loads in 2011 (Fig.~\ref{fig:fig4}).

@@

{\centering \includegraphics[width=0.5\linewidth,]{Schramm-2023-06-PeerJ_files/figure-latex/fig4-1}

@@

{\centering \includegraphics[width=1\linewidth,]{Schramm-2023-06-PeerJ_files/figure-latex/fig5-1}

@@

{\centering \includegraphics[width=1\linewidth,]{Schramm-2023-06-PeerJ_files/figure-latex/fig6-1}

@@

{\centering \includegraphics[width=1\linewidth,]{Schramm-2023-06-PeerJ_files/figure-latex/fig7-1}

@@

{\centering \includegraphics[width=1\linewidth,]{../fig8}

@@

The non-linear estuary water quality trends identified in the current
study differed slightly from previously identified trends
\autocite{bugica_water_2020}. This is not unexpected due to the
different time periods, different methodology, and generally small
slopes identified for most of the significant water quality parameters
in prior work. Both DO and cholorophyll-\emph{a} concentrations at all
three Lavaca Bay sites were stable from 2005 through 2020. This is a
positive outcome in comparison to other Texas estuaries that are facing
larger demands for freshwater diversions, higher population growth, and
more intense agricultural production which have resulted in more direct
signs of eutrophication
\autocite{wetzWaterQualityDynamics2016,bugica_water_2020}. Despite the
stability of DO and cholorophyll-\emph{a}, there are concerning site
specific increases in NO\textsubscript{\emph{X}} and TKN concentration
over the same time period. These trends are espcially concerning due to
the nitrogen limitation identified in many Texas estuaries
\autocite{gardnerNitrogenFixationDissimilatory2006,houTransformationFateNitrate2012,doradoUnderstandingInteractionsFreshwater2015,paudelRelationshipSuspendedSolids2019,wetz_exceptionally_2017}
and the relatively low ambient concentrations observed in Lavaca Bay.

The strong positive effect of freshwater inflow on
NO\textsubscript{\emph{X}}, TKN, and TP concentration are suggestive of
nonpoint watershed sources, consistent with watershed uses and with
other studies relating freshwater inflow with nutrient concentrations in
Lavaca Bay and other estuaries
\autocite{russell_effect_2006,caffreyHighNutrientPulses2007,peierlsNonmonotonicResponsesPhytoplankton2012,palmerImpactsDroughtsLow2015,ciraPhytoplanktonDynamicsLowinflow2021}.
Inflow had a non-linear relationship with TKN at the two upstream sites,
with TKN increasing as freshwater inflow transitioned from low to
moderate levels. At higher freshwater inflows, the effect was
attenuated, possibly indicating a flushing effect at higher freshwater
inflow. No relationship between TKN and freshwater inflow were observed
at TCEQ-13384 located in the lower reach of Lavaca Bay. The results
suggest that tidal flushing maybe diluting TKN and act as a control on
the effects of freshwater inflow in the lower reaches of Lavaca Bay. The
results are also consistent with previous work that suggest processing
of organic loads in the upper Lavaca Bay or tidal portions of the Lavaca
River reduce transport of nutrients to the lower reaches of Lavaca Bay
\autocite{russell_effect_2006}.

@@

The primary purpose of this study was to provide estimates of watershed
nutrient loadings and assess water quality responses to changes in
nutrient loads. GAMs provided reliable estimates of watershed
NO\textsubscript{3} and TP loads. However, additional flow-biased data
collection efforts is needed to decrease the prediction variance and
improve accuracy at critical high-flow loading events. While some
ongoing projects will fill data gaps for total nitrogen and TKN loading,
additional efforts are needed to coordinate data collection efforts
specifically for load estimation across Texas estuaries. Despite these
data gaps, this study identified high annual fluctuations in nutrient
loads driven primarily by discharge. No evidence was identified to
indicate that on the ground management had changed nutrient loading in
the Navidad River subwatershed. There was weak evidence for recent
reductions in flow-normalized NO\textsubscript{3} loading in the Lavaca
River subwatershed although the results are at odds with flow-normalized
trends in TP loads.

>>>>

\begin{abstract}
Evaluating estuary water quality responses to reductions (or increases)
in nutrient loading attributed to on the ground management actions can
be challenging due to the strong influence of environmental drivers on
nutrient loads and non-linear relationships. This study applied
semiparametric approaches to calculate watershed nutrient loads and
assess responses in estuary water quality to seasonally-adjusted
freshwater inflow and flow-adjusted nutrient loads in Lavaca Bay, Texas.
Lavaca Bay is a secondary embayment on the Texas coast displaying early
potential for eutrophication and water quality degradation. Use of
flow-adjusted nutrient loads allowed the study to evaluate the response
in water quality to changes in nutrient loads driven by non-natural
variation. Cross-validation indicated that, despite data constraints,
semiparametric models performed well at nutrient load prediction. Based
on these models, delivered annual nutrient loads varied substantially
from year to year. In contrast, minimal changes in flow-normalized loads
indicate that nutrient loadings were driven by natural variation in
precipitation and runoff as opposed to changes in management of nonpoint
sources. Models indicated no evidence of long-term changes in dissolved
oxygen or chlorophyll-\emph{a} within Lavaca Bay. However, site specific
long-term increases in both organic and inorganic nitrogen are
concerning for their potential to fuel eutrophication. Further analysis
found freshwater inflow had strong influences on nutrient and
chlorophyll-\emph{a} concentrations but there was no evidence that
changes in watershed nutrient loading explained additional variation in
dissolved oxygen or chlorophyll-\emph{a} concentrations. In addition to
providing a baseline assessment of watershed nutrient loading and water
quality responses in the Lavaca Bay watershed, this study provides
methodological support for the use of semiparametric models in load
regression models and estuary assessments.
% Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text.
\end{abstract}

@@

Similar to many coastal areas globally, the coastal watersheds along the
Texas Gulf coast are facing pressures from growing populations,
increases in point source and non-point source pollution and alterations
to freshwater flows that degrade water quality in downstream estuaries
\autocite{bricker_effects_2008,kennicuttWaterQualityGulf2017,bugica_water_2020}.
Despite these escalating pressures, national scale assessments have
classified coastal estuaries in Texas as moderate or low risk for
eutrophic conditions \autocite{bricker_effects_2008}. However, a suite
of recent studies indicates that estuary water quality dynamics in both
agricultural and urban dominated watersheds within Texas are expressing
conditions that are increasingly conducive to algal blooms and
eutrophication
\autocite{wetzWaterQualityDynamics2016,wetz_exceptionally_2017,bugica_water_2020,chinPhytoplanktonBiomassCommunity2022}.
With identification of several localized areas of estuary water quality
concern along the Texas coast \autocite{bugica_water_2020}, localized
studies are being prioritized to better inform management actions.

The goal of this project is to assess watershed nutrient loading and the
resulting water quality responses in Lavaca Bay, Texas. Lavaca Bay is a
secondary bay in the larger Matagorda Bay system located roughly halfway
between Houston, Texas and Corpus Christi, Texas. Lavaca Bay faces
substantial challenges associated with legacy contamination but general
water quality parameters such as dissolved oxygen (DO), nutrients, and
biological parameters have been well within state water quality
standards. Despite largely meeting state designated water quality
thresholds, there have been concerning declines in abundance, biomass,
and diversity of benthic fauna in Lavaca Bay
\autocite{beserespollackLongtermTrendsResponse2011}. These declines are
partially attributed to reductions in freshwater inflow and changes in
estuary salinity and are indicative of an already stressed system
\autocite{beserespollackLongtermTrendsResponse2011,palmerImpactsDroughtsLow2015,montagnaAssessmentRelationshipFreshwater2020}.
More recently, significant linear increases in total phosphorus (TP),
orthophosphate, total Kjeldahl nitrogen (TKN), and chlorophyll-\emph{a}
concentrations were identified at monitoring sites within Lavaca Bay
\autocite{bugica_water_2020}. These trends in nutrient concentrations
are concerning due to the role of nitrogen as a limiting factor for
primary production in many Texas estuaries
\autocite{gardnerNitrogenFixationDissimilatory2006,houTransformationFateNitrate2012,doradoUnderstandingInteractionsFreshwater2015,paudelRelationshipSuspendedSolids2019,wetz_exceptionally_2017}
and the ramifications that changes in nitrogen loadings could have for
productivity and eutrophication in Lavaca Bay.

There are ongoing efforts between local, state, and federal agencies to
address water quality impairments in the freshwater portions of the
Lavaca Bay watershed
\autocite{jainTechnicalSupportDocument2021,schrammLavacaRiverWatershed2018,bertholdDirectMailingEducation2021}.
However, on a statewide scale, these approaches have shown limited
success and emphasize a need for improved methods to assess and link
management actions with downstream water quality
\autocite{schrammTotalMaximumDaily2022}.\\
Such methods could help identify and replicate effective water quality
management actions across the state . The identification and
communication of changes and trends in water quality is complicated by
the fact that trends are often non-linear and confounded by
precipitation and runoff that hinder traditional analysis
\autocite{wazniakLinkingWaterQuality2007,lloydMethodsDetectingChange2014}.
The development and application of flexible statistical methods such as
Weighted Regressions on Time, Discharge and Season
\autocite[WRTDS,][]{hirschWeightedRegressionsTime2010} and Generalized
Additive Models \autocite[GAMs,][]{woodFastStableRestricted2011} have
provided effective tools for researchers to quantify and communicate
non-linear changes in river and estuary pollutant loadings.

WRTDS calculates a time series of in-stream concentrations or loads
(daily, monthly, or annually) and flow-normalized estimates of
concentrations and loads using locally weighted regression for unique
combinations of time, discharge, and season. WRTDS has been widely used
to assess and identify trends in riverine nutrients
\autocite{oelsner_recent_2019,stackpooleLongTermMississippi2021},
chlorides \autocite{stetsIncreasingChlorideRivers2018}, and other
pollutants of concern \autocite{shodaWaterqualityTrendsRivers2019}.
WRTDS has also been successfully adapted to assess trends in estuary
water quality concentrations \autocite{beckFourDecadesWater2018}.

While WRTDS is a statistical approach developed specifically for water
quality applications, GAMs are a broadly applicable statistical method.
GAMs are a semiparametric extension of generalized linear models where
the linear predictor is represented as the sum of multiple unknown
smooth functions and parametric linear predictors
\autocite{woodFastStableRestricted2011}. Although the underlying
parameter estimation procedure of GAMs are substantially different than
WRTDS, both the functional form and results have been demonstrated to be
similar when assessing nutrient concentration trends
\autocite{beckNumericalQualitativeContrasts2017}. Water quality
applications of GAMs have included river and catchment nutrient
concentration and load estimation
\autocite{wangLoadEstimationUncertainties2011,kroonRiverLoadsSuspended2012,kuhnertQuantifyingTotalSuspended2012,robson_prediction_2015-1,hagemannEstimatingNutrientOrganic2016,mcdowell_implications_2021,biagi_novel_2022},
and assessment of temporal trends of nutrients
\autocite{beckNumericalQualitativeContrasts2017,murphyGeneralizedAdditiveModel2019},
phytoplankton \autocite{bergbuschUnexpectedShiftPhytoplankton2021}, and
cyanobacteria \autocite{hayesEffectsLakeWarming2020}. Recently GAMs have
also been used to link water quality responses in receiving water bodies
to changes in nonpoint source nutrient inputs
\autocite{murphyNutrientImprovementsChesapeake2022}. For a substantial
discussion on the differences (and similarities) between GAMs and WRTDS
for water quality applications readers are referred to
\textcite{beckNumericalQualitativeContrasts2017}.

@@

{\centering \includegraphics[width=5.2in,]{Schramm-2023-06-PeerJ_files/figure-latex/fig1}

@@

Water quality data for the two freshwater and three estuary locations
were obtained from the Texas Commission on Environmental Quality (TCEQ)
Surface Water Quality Monitoring Information System. Data submitted
through the system are required to be collected under Quality Assurance
Project Plans and lab method procedures outlined by the TCEQ's
procedures manual. These operating procedures ensure consistent
collection and laboratory methods are applied between samples collected
by different entities and under different projects. All sites had
varying time periods and availability of water quality data. For
freshwater locations, TP from January 2000 through December 2020 and
nitrate-nitrogen (NO\textsubscript{3}) data from January 2005 through
December 2020 were downloaded (Table~\ref{tab:table1}). Less than
5-years of total nitrogen and TKN concentration data were available at
the freshwater sites and deemed insufficient to develop load estimation
models
\autocite{horowitzEvaluationSedimentRating2003,snelderEstimationCatchmentNutrient2017}.
The three estuary sites included an upper Lavaca Bay site near the
outlet of the Lavaca River system (TCEQ-13563), a mid-Lavaca Bay site
(TCEQ-13383), and the lower Lavaca Bay site near the mouth of the Bay
(TCEQ-13384). For estuary locations, we obtained data for TP,
Nitrite+Nitrate (NO\textsubscript{\emph{X}}), TKN, chlorophyll-\emph{a},
and DO concentrations from January 2005 through December 2020
(Table~\ref{tab:table2}).

@@

\begin{align}
g(\mu) &= \alpha + f_1(ddate) + f_2(yday) + f_3(log(1+Q)) + f_4(ma) + f_5(fa), \nonumber \\
y &\sim \mathcal{N}(\mu,\,\sigma^{2})
\label{eq1}
\end{align}

where \(\mu\) is the conditional expected NO\textsubscript{3} or TP
concentration, \(g()\) is the log-link, \(\alpha\) is the intercept,
\(f_n()\) are smoothing functions. \(y\) is the response variable
(NO\textsubscript{3} or TP concentration) modeled as normally
distributed with mean \(\mu\) and standard deviation \(\sigma\).
\emph{ddate} is the date converted to decimal notation, \emph{yday} is
numeric day of year (1-366), and \emph{Q} is the mean daily streamflow.

@@

Daily loads were estimated as the predicted concentration multiplied by
the daily streamflow. For the Navidad River (USGS-08164525) site, daily
loads at the dam were calculated from the discrete daily concentration
at the discharge point of the lake and corresponding reported daily
discharge from the dam. Flow-normalized loads were estimated similar to
WRTDS by setting flow-based covariates on each day of the year equal to
each of the historical values for that day of the year over the study
period \autocite{hirschWeightedRegressionsTime2010}. The flow-normalized
estimate was calculated as the mean of all the predictions for each day
considering all possible flow values. Standard deviations and 90\%
credible intervals were obtained by drawing samples from the
multivariate normal posterior distribution of the fitted GAM
\autocite{woodConfidenceIntervalsGeneralized2006,marraCoveragePropertiesConfidence2012,mcdowell_implications_2021}.
GAM performance was evaluated with repeated 5-fold cross validation
\autocite{burmanComparativeStudyOrdinary1989} using average
Nash-Sutcliffe Efficiency (NSE), Pearson sample correlation (\emph{r})
and percent bias (PBIAS) metrics.

@@

Because streamflow and nutrient loads are tightly correlated, freshwater
inflow can mask signals due to changes in nutrient loads alone.
Following the methodology implemented by
\textcite{murphyNutrientImprovementsChesapeake2022}, both freshwater
inflow and nutrient loads were prepossessed to account for season and
streamflow respectively. Raw freshwater inflow values were replaced by
seasonally adjusted log transformed inflow obtained from the residuals
of a GAM model fit between season(day of year) and log transformed daily
freshwater inflow. Raw nutrient loads were replaced with flow-adjusted
values obtained from the residuals of a GAM model relating log
transformed NO\textsubscript{3} or TP loads to log transformed daily
inflow. Response residuals from the respective GAM models were used as
\emph{Q} and \emph{Load} in Eq.~\ref{eq6} and Eq.~\ref{eq7}.

@@

Predictive performance of nutrient loads ranged from ``satisfactory'' to
``very good'' based on standardized evaluation metrics of NSE, \emph{r},
and PBIAS \autocite{moriasiHydrologicWaterQuality2015} calculated using
5-fold cross validation. Median goodness-of-fit metrics for
NO\textsubscript{3} models in the Lavaca River were 0.34 NSE, 0.70
\emph{r}, and 2.00 PBIAS. Navidad River NO\textsubscript{3} models
appeared to perform slightly better with 0.48 NSE and 0.87 \emph{r} but
with higher bias at 10.90 PBIAS. Generally, TP models performed better
than NO\textsubscript{3} models. Median goodness-of-fit metrics for TP
in the Lavaca River were 0.81 NSE, 0.93 \emph{r}, and -7.20 PBIAS.
Navidad River TP models had similar performance with 0.91 NSE, 0.99
\emph{r}, and -3.30 PBIAS. Density plots of metrics show similar
distribution of values between sites for the same parameter, with the
exception \emph{r} values for NO\textsubscript{3} loads where Lavaca
River had a much larger variance in values compared to the Navidad River
(Fig.~\ref{fig:fig2}). TP GAMS had higher average NSE and \emph{r}
values and lower variance in metric values compared to
NO\textsubscript{3}.

@@

{\centering \includegraphics[width=1\linewidth,]{Schramm-2023-06-PeerJ_files/figure-latex/fig2}

@@

{\centering \includegraphics[width=1\linewidth,]{Schramm-2023-06-PeerJ_files/figure-latex/fig3}

@@

Aggregated across both sites, the mean annual NO\textsubscript{3} load
from 2005 through 2020 was 205,405 kg (126,867 kg - 341,569 kg, 90\%
CI). Annual NO\textsubscript{3} loads ranged from 12,574 kg in 2011 to
794,510 kg in 2007. Total annual TP loads ranged from 7,839 kg in 2011
to 595,075 kg in 2007. Mean annual TP loading from 2005 through 2020 was
182,673 kg (152,227 kg - 219,310 kg, 90\% CI). On average, the Navidad
River accounted for 68\% of NO\textsubscript{3} loads and 59\% of TP
loads from 2005 through 2020. However, during periods of extreme drought
the Lavaca River became the primary source of nutrient loading in the
watershed with the Navidad River only accounting for 15\% and 25\% of
NO\textsubscript{3} and TP loads in 2011 (Fig.~\ref{fig:fig4}).

@@

{\centering \includegraphics[width=0.5\linewidth,]{Schramm-2023-06-PeerJ_files/figure-latex/fig4}

@@

{\centering \includegraphics[width=1\linewidth,]{Schramm-2023-06-PeerJ_files/figure-latex/fig5}

@@

{\centering \includegraphics[width=1\linewidth,]{Schramm-2023-06-PeerJ_files/figure-latex/fig6}

@@

{\centering \includegraphics[width=1\linewidth,]{Schramm-2023-06-PeerJ_files/figure-latex/fig7}

@@

{\centering \includegraphics[width=1\linewidth,]{Schramm-2023-06-PeerJ_files/figure-latex/fig8}

@@

The non-linear estuary water quality trends identified in the current
study differed slightly from previously identified trends
\autocite{bugica_water_2020}. This is not unexpected due to the
different time periods, different methodology, and generally small
slopes identified for most of the significant water quality parameters
in prior work. Both DO and cholorophyll-\emph{a} concentrations at all
three Lavaca Bay sites were stable from 2005 through 2020. This is a
positive outcome in comparison to other Texas estuaries that are facing
larger demands for freshwater diversions, higher population growth, and
more intense agricultural production which have resulted in more direct
signs of eutrophication
\autocite{wetzWaterQualityDynamics2016,bugica_water_2020}. Despite the
stability of DO and cholorophyll-\emph{a}, there are concerning site
specific increases in NO\textsubscript{\emph{X}} and TKN concentration
over the same time period. These trends are especially concerning due to
the nitrogen limitation identified in many Texas estuaries
\autocite{gardnerNitrogenFixationDissimilatory2006,houTransformationFateNitrate2012,doradoUnderstandingInteractionsFreshwater2015,paudelRelationshipSuspendedSolids2019,wetz_exceptionally_2017}
and the relatively low ambient concentrations observed in Lavaca Bay.

The strong positive effect of freshwater inflow on
NO\textsubscript{\emph{X}}, TKN, and TP concentration are suggestive of
nonpoint watershed sources, consistent with watershed uses and with
other studies relating freshwater inflow with nutrient concentrations in
Lavaca Bay and other estuaries
\autocite{russell_effect_2006,caffreyHighNutrientPulses2007,peierlsNonmonotonicResponsesPhytoplankton2012,palmerImpactsDroughtsLow2015,ciraPhytoplanktonDynamicsLowinflow2021}.
Inflow had a non-linear relationship with TKN at the two upstream sites,
with TKN increasing as freshwater inflow transitioned from low to
moderate levels. At higher freshwater inflows, the effect was
attenuated, possibly indicating a flushing effect at higher freshwater
inflow. No relationship between TKN and freshwater inflow were observed
at TCEQ-13384 located in the lower reach of Lavaca Bay. The results
suggest that tidal flushing might dilute TKN and act as a control on the
effects of freshwater inflow in the lower reaches of Lavaca Bay. The
results are also consistent with previous work that suggest processing
of organic loads in the upper Lavaca Bay or tidal portions of the Lavaca
River reduce transport of nutrients to the lower reaches of Lavaca Bay
\autocite{russell_effect_2006}.

@@

The primary purpose of this study was to provide estimates of watershed
nutrient loadings and assess water quality responses to changes in
nutrient loads. GAMs provided reliable estimates of watershed
NO\textsubscript{3} and TP loads. However, additional flow-biased data
collection efforts are needed to decrease the prediction variance and
improve accuracy at critical high-flow loading events. While some
ongoing projects will fill data gaps for total nitrogen and TKN loading,
additional efforts are needed to coordinate data collection efforts
specifically for load estimation across Texas estuaries. Despite these
data gaps, this study identified high annual fluctuations in nutrient
loads driven primarily by discharge. No evidence was identified to
indicate that on the ground management had changed nutrient loading in
the Navidad River subwatershed. There was weak evidence for recent
reductions in flow-normalized NO\textsubscript{3} loading in the Lavaca
River subwatershed although the results are at odds with flow-normalized
trends in TP loads.
